A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-04 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.tex
COVID-19 Municipios v0

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

>>>>

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.\\
  \end{abstract}

### 2020-05-04 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
COVID-19 Municipios v0

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~(2020), examined data from China and came to
the conclusion that neither temperature nor ultraviolet indices had an
association with transmission of COVID-19. This is despite previous
research that has linked less exposure to UVB radiation to higher
prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
independent varaibles on \(y_t\). This flexibility can be reduced and it
is posible to impose restrinctions considering some \(\beta\)
coefficients as being constant over time. In this case, we can
reformulate the coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

@@

\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
\centering
\begin{tabular}[t]{lrrrl}
\toprule
Variable & Min & Mean & Max & Note\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Direct Effects}}\\
\hspace{1em}log(GDPpc) & 0.675 & 0.675 & 0.675 & Constrained\\
\hspace{1em}log(Older) & -0.802 & -0.802 & -0.802 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.167 & Non-constrained\\
\hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.562 & -0.544 & -0.031 & Non-constrained\\
\hspace{1em}log(Temperature) & -2.019 & -1.421 & -0.929 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
\hspace{1em}log(GDPpc) & 0.638 & 0.638 & 0.638 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.758 & -0.758 & -0.758 & Constrained\\
\hspace{1em}log(Density) & -0.032 & -0.004 & 0.157 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.341 & Non-constrained\\
\hspace{1em}log(Humidity) & -1.475 & -0.157 & -0.001 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.644 & -0.274 & -0.014 & Non-constrained\\
\hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Total Effects}}\\
\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
\bottomrule
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}
\end{table}

>>>>

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~(2020), examined data from China and came to
the conclusion that neither temperature nor ultraviolet indices had an
association with transmission of COVID-19. This is despite previous
research that has linked less exposure to UVB radiation to higher
prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
the independent variables on \(y_t\). This flexibility can be reduced
and it is posible to impose restrictions considering some \(\beta\)
coefficients as being constant over time. In this case, we can
reformulate the coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

@@

\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
\centering
\begin{tabular}[t]{lrrrl}
\toprule
Variable & Min & Mean & Max & Note\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Direct Effects}}\\
\hspace{1em}log(GDPpc) & 0.674 & 0.674 & 0.674 & Constrained\\
\hspace{1em}log(Older) & -0.801 & -0.801 & -0.801 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.166 & Non-constrained\\
\hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.559 & -0.544 & -0.031 & Non-constrained\\
\hspace{1em}log(Temperature) & -2.019 & -1.420 & -0.929 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
\hspace{1em}log(GDPpc) & 0.639 & 0.639 & 0.639 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.759 & -0.759 & -0.759 & Constrained\\
\hspace{1em}log(Density) & -0.032 & -0.004 & 0.158 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.342 & Non-constrained\\
\hspace{1em}log(Humidity) & -1.478 & -0.157 & -0.001 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.647 & -0.275 & -0.014 & Non-constrained\\
\hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Total Effects}}\\
\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
\bottomrule
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}
\end{table}
